---
title: "Illinois2018"
author: "Helen Miller - modified by Martin Menninger"
date: "July 11, 2017 --> 7/25/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
library(MazamaSpatialUtils)
library(sp)
library(rgdal)
library(purrr)
library(animation)
library(ggplot2)
library(tidyverse)
library(dbplyr)
library(ggthemes)

source("~/R/NBI/R/convert2016_CMAP_2.R")

#nbi <- convert2016("~/R/NBI/Data/nbiIL.rds")

#list[nbi, nbi_trend] <-  convert2016("~/R/NBI/Data/Raw/nbiIL92_18.RDS")
nbi_trend <-  convert2016("~/R/NBI/Data/Raw/nbiIL92_18.RDS")
nbi <- nbi_trend %>% filter(NBI_Year == max(nbi_trend$NBI_Year))


setSpatialDataDir("~/R/NBI/Data/Spatial")
#MazamaSpatialUtils::convertUSCensusCounties()
#MazamaSpatialUtils::convertUSCensusStates()
loadSpatialData("USCensusCounties")
loadSpatialData("USCensusStates")
conus <- subset(USCensusStates, !stateCode %in% c("HI", "AK","GU","AS", "PR", "HI", "VI", "MP"))
conus <- rmapshaper::ms_simplify(conus, .05)
nbisp <- subset(nbi, is.na(latitude) == F)
nbisp <- subset(nbisp, is.na(longitude) == F)
coordinates(nbisp) <- c("longitude", "latitude")
proj4string(nbisp) <- proj4string(USCensusStates)
#il <- subset(nbisp, state_code == "17")
ilCounties <- subset(USCensusCounties, stateCode == "IL")
nbi$county <- getUSCounty(nbi$longitude, nbi$latitude, dataset = "ilCounties", stateCode = "IL")
```

 What is so unique about Illinois?

Illinois is an interesting case for a couple of reasons. Bridges are not equally distributed throughout the country. There are a lot more bridges in the Northeast of the state. Fewer in the less-developed south. (This map does not really work anymore.)

```{r conusBridges}
plot(spTransform(conus, CRS("+init=epsg:5071")))
plot(subset(spTransform(conus, CRS("+init=epsg:5071")), stateCode == "IL"), border = "red", add = TRUE, lwd = 5)
points(spTransform(nbisp, CRS("+init=epsg:5071")), pch = 17, cex = .05)
title("Bridges in the Continental US")
```
  
In Illinois, like most states, the peak of bridge building was in the 60's or near that time. These bridges are now 60 years old, and many are nearing the end of their useful life. 


    
# Overall Distribution of Bridges in Illinois
    
```{r USBridgeMap}
plot(subset(conus, stateCode == "IL"))
#points(nbi$longitude, nbi$latitude, col = "red", cex = .1)
points(nbi$longitude[nbi$water==1], nbi$latitude[nbi$water==1], col = "blue", cex = .09)
points(nbi$longitude[nbi$water==0], nbi$latitude[nbi$water==0], col = "red", cex = .1)
#points(subset(nbi, water == 1), pch = 2, cex = .1, col = "blue")
#points(subset(nbi), water == 0), pch = 2, cex = .3, col = "red")
title("All bridges in Illinois")
legend("bottomleft", c("Bridges over water", "Bridges not over water"),
       pch = 17, col = c("blue", "red"), title = "Legend")
```



```{r CMAPBridgeMap}
plot(subset(ilCounties, countyFIPS %in% c("031", "043", "089", "093", "097", "111", "197")))
points(nbi$longitude[nbi$water==1], nbi$latitude[nbi$water==1], col = "blue", cex = .4)
points(nbi$longitude[nbi$water==0], nbi$latitude[nbi$water==0], col = "red", cex = .5)
title("All CMAP bridges")
legend("bottomleft", c("Bridges over water", "Bridges not over water"),
       pch = 17, col = c("blue", "red"), title = "Legend")
```


# Bridge Age  
  
```{r yearBuiltHistogram}
hist(nbi$yearBuilt, xlab = "Year Built", main = "Year Built", n = 100)
```

```{r yearBuiltGgplot}

#ilBridges <- as.data.frame(il)

ggplot(nbi,  aes(x=yearBuilt))+
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(1870, 2020, 20), limits = c(1870, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "Year Illinois Bridges Built")

ggplot(nbi,  aes(x=yearBuilt, fill = Description))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1880, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "Year All Illinois Bridges Built")

nbi %>%
  filter(OWNER_022 == "31") %>%
ggplot(aes(x=yearBuilt, fill = OWNER_022))+
  geom_bar(position = "stack", show.legend = FALSE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1880, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "Year Tollway Bridges Built")

nbi %>%
  filter(OWNER_022 == "01") %>%
ggplot(aes(x=yearBuilt, fill = Description))+
  geom_bar(position = "stack", show.legend = FALSE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1880, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "Year State Bridges Built")

ggplot(nbi,  aes(x=yearBuilt, fill = FUNCTIONAL_CLASS_026))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1880, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "Year All Illinois Bridges Built")

ggplot(subset(nbi, FUNCTIONAL_CLASS_026 %in% (c("01", "11"))),  aes(x=yearBuilt, fill = FUNCTIONAL_CLASS_026))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1930, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "IL Interstates by Year Built (01= Rural, 11 = Urban)")

```


  
With so many bridges destroyed, we would expect that just as many bridges had to be rebuilt. Thus, is the huge spike in bridge building in 1935 a sign of enthusiastic expansion of road networks, an indication of the aftermath of a natural disaster which destroyed many bridges, or simply an artifact of human fallibility in recording data? Let's take a closer look at the bridges built in 1935.  
  
```{r floodBridges}
floodBridges <- subset(nbi, yearBuilt == 1958)
```

  
According to our dataset, `r nrow(floodBridges)` were built in 1935. That's a lot. 307 highway bridges were reportedly destroyed in the flood, and, we expect, rebuilt shortly thereafter. This does not account for railroad bridges or bridges on small county roads, so it is fair to expect the total number of reconstructed bridges to be significantly higher than 307. Is it plausible that the vast majority of bridges built in 1935 were non-highway bridges destroyed in the flood? They do seem to be primarily non-highway bridges. The average daily traffic for bridges built in 1935 is $`r mean(floodBridges$ADT_029)`$ vehicles where as the average daily traffic for all Illinois bridges is $`r mean(nbi$ADT_029)`$ vehicles. The vast majority of the bridges built in 1935 carry 100 cars or fewer. In fact, only 11% of the bridges built in 1935 carry over 100 cars per day, on average, which indicates that they are generally small, non-highway bridges, the kind that would easily get washed out in a flood. However, it is also likely that most large bridges were built more recently, once major highways were developed.

```{r traffic1935, fig.width=7}
hist(floodBridges$ADT_029, n = 100000, main = "Traffic on Bridges Built in 1958", xlab = "Average Daily Traffic")
hist(floodBridges$ADT_029, n = 100000, xlim = c(0,500), main = "Zoomed-in Histogram of Traffic on Bridges Built in 1958", xlab = "Average Daily Traffic")
```


Now, let's take a look at where these bridges actually are.  
  
```{r bridges1935Map}
plot(subset(USCensusStates, stateCode == "IL"))
points(floodBridges$longitude, floodBridges$latitude, pch = 17, cex = .5)
title("Bridges built in 1958")
```
  
The Republican River winds through Southern Nebraska. Many of the 1935 bridges were, indeed, built near the Republican River. However, just as many, if not more, were built in other parts of Nebraska. It seems most likely that the huge spike in bridge building was more due to something like age estimations by surveyors which defaulted to 1935.



```{r bridgeAgeByCountyMap, eval = FALSE}
breaks <- c(0,seq(16,136, by = 20))
ilColors <- RColorBrewer::brewer.pal(7, "PuRd")
colorIndex <- .bincode(nbi$age, breaks = breaks)
nbi$colorIndex <- colorIndex
plot(subset(USCensusStates, stateCode == "IL"))
points(nbi$longitude, nbi$latitude, cex = .3, pch = 17, col = ilColors[colorIndex])
legend("bottomleft", c("1880-1900", "1900-1920", "1920-1940", "1940-1960", "1960-1980", "1980-2000", "2000-2017"), pch = 17, col = ilColors[7:1])
```
```{r bridgeAgeByCountyMapCMAP, eval = FALSE}
breaks <- c(0,seq(16,136, by = 20))
ilColors <- RColorBrewer::brewer.pal(7, "PuRd")
colorIndex <- .bincode(nbi$age, breaks = breaks)
nbi$colorIndex <- colorIndex
plot(subset(ilCounties, countyFIPS %in% c("031", "043", "089", "093", "097", "111", "197")))
points(nbi$longitude, nbi$latitude, cex = .3, pch = 17, col = ilColors[colorIndex])
legend("bottomleft", c("1880-1900", "1900-1920", "1920-1940", "1940-1960", "1960-1980", "1980-2000", "2000-2017"), pch = 17, col = ilColors[7:1])
```


```{r highwayBridgeMap, include = FALSE}
breaks <- c(0,seq(16,136, by = 20))
ilColors <- RColorBrewer::brewer.pal(7, "PuRd")
colorIndex <- .bincode(nbi$age, breaks = breaks)
breaks <- c(0,seq(16,136, by = 20))
plot(subset(USCensusStates, stateCode == "IL"))
points(nbi$longitude, nbi$latitude, cex = .5, pch = 17, col = ilColors[colorIndex])
```
  
# Traffic


```{r trafficMap}
par(mar = c(4,0,4,0))
layout(matrix(c(2,1),1,2), widths = c(2,8))
plot(subset(USCensusStates, stateCode == "IL"))
points(nbi$longitude, nbi$latitude, pch = 2, cex = nbi$ADT_029/80000)
title("Traffic on Illinois's Bridges")
plot.new()
legend("left", c("1,000", "10,000", "100,000"), pch = 2, pt.cex = c(1000/20000, 10000/20000, 100000/20000), title = "Daily motor Traffic", y.intersp = 2)
```

```{r trafficMapCMAP}
par(mar = c(4,0,4,0))
layout(matrix(c(2,1),1,2), widths = c(2,8))
plot(subset(ilCounties, countyFIPS %in% c("031", "043", "089", "093", "097", "111", "197")))
points(nbi$longitude, nbi$latitude, pch = 2, cex = nbi$ADT_029/80000)
title("Traffic on CMAP's Bridges")
plot.new()
legend("left", c("1,000", "10,000", "100,000"), pch = 2, pt.cex = c(1000/80000, 10000/80000, 100000/80000), title = "Daily motor Traffic", y.intersp = 2)
```

```{r logTrafficMap}
trafficColors <- RColorBrewer::brewer.pal(9,"YlOrRd")
breaks <- c(seq(0, log(max(nbi$ADT_029)), length = 10))
colorIndex <- .bincode(log(nbi$ADT_029), breaks = breaks)
plot(subset(USCensusStates, stateCode == "IL"))
points(nbi$longitude, nbi$latitude, pch = 17, col = trafficColors[colorIndex], cex = log(nbi$ADT_029)/20)
legend("bottomleft", c('100','1000','10000','100000'), col = trafficColors[c(4,6,7,9)], pch = 17, pt.cex = log(c(100,1000,10000,100000))/20, title = "Daily Motor Traffic")
title("Traffic on Illinois's Bridges")
```

```{r logTrafficMapCMAP}
trafficColors <- RColorBrewer::brewer.pal(9,"YlOrRd")
breaks <- c(seq(0, log(max(nbi$ADT_029)), length = 10))
colorIndex <- .bincode(log(nbi$ADT_029), breaks = breaks)
plot(subset(ilCounties, countyFIPS %in% c("031", "043", "089", "093", "097", "111", "197")))
points(nbi$longitude, nbi$latitude, pch = 17, col = trafficColors[colorIndex], cex = log(nbi$ADT_029)/20)
legend("bottomleft", c('100','1000','10000','100000'), col = trafficColors[c(4,6,7,9)], pch = 17, pt.cex = log(c(100,1000,10000,100000))/20, title = "Daily Motor Traffic")
title("Traffic on CMAP's Bridges")
```
As expected, traffic is heaviest along the highways. There are particular high-traffic clumps at Omaha and Lincoln. Most bridges have low traffic. Less than about 1,000 vehicles per day and they barely show up on the map. 
  
  
# Aggregated by county: 

To see how bridge age is distributed geographically, let's aggregate the data by county.
  
```{r meanAgeByCountyMap}
meanAgeByCounty <- aggregate(data=nbi, age~county, FUN = mean)
breaks <- c(24.4, seq(30,55, by = 5), 60.2)
colorIndex <- .bincode(meanAgeByCounty$age, breaks = breaks)
names(colorIndex) <- meanAgeByCounty$county
plot(ilCounties, col = ilColors[colorIndex[ilCounties$countyName]])
legend("bottomleft", c("25-30", "30-35", "35-40", "40-45", "45-50","50-55", "55-60"), pch = 15, col = ilColors, title = "Mean Bridge Age")
title("Mean bridge age by county")
```

No clear patterns arise when we look at the mean age by county. Let's take a look at the mode year which bridges were built in each county.



```{r modeAgeByCountyMap}
getmode <- function(v) {
uniqv <- unique(v)
uniqv[which.max(tabulate(match(v, uniqv)))]
}

modeYearByCounty <- aggregate(data = nbi, yearBuilt~county, FUN = getmode)
hist(modeYearByCounty$yearBuilt)
breaks <- c(seq(1929, 1999, by = 10), 2017)
modeIndex <- .bincode(modeYearByCounty$yearBuilt, breaks = breaks)
modeColors <- RColorBrewer::brewer.pal(8, "PuRd")
names(modeIndex) <- modeYearByCounty$county
plot(ilCounties, col = modeColors[8:1][modeIndex[ilCounties$countyName]])
legend("bottomleft", c("1930s", "1940s", "1950s", "1960s", "1970s", "1980s", "1990s", "2000s"), pch = 15, col = modeColors[8:1])
points(nbi, pch = 17, cex = .2, col = "green")
title("Mode year bridges were built")
```


```{r modeAgeByCountyMapCMAP}
getmode <- function(v) {
uniqv <- unique(v)
uniqv[which.max(tabulate(match(v, uniqv)))]
}

modeYearByCounty <- aggregate(data = nbi, yearBuilt~county, FUN = getmode)
hist(modeYearByCounty$yearBuilt)
breaks <- c(seq(1929, 1999, by = 10), 2017)
modeIndex <- .bincode(modeYearByCounty$yearBuilt, breaks = breaks)
modeColors <- RColorBrewer::brewer.pal(8, "PuRd")
names(modeIndex) <- modeYearByCounty$county

plot(subset(ilCounties, countyFIPS %in% c("031", "043", "089", "093", "097", "111", "197")), col = modeColors[8:1][modeIndex[ilCounties$countyName]])
legend("bottomleft", c("1930s", "1940s", "1950s", "1960s", "1970s", "1980s", "1990s", "2000s"), pch = 15, col = modeColors[8:1])
points(nbi, pch = 17, cex = .2, col = "green")
title("Mode year bridges were built")
```


As expected, in most counties, most bridges were built in the 1930s. There is a group of counties near the middle of the state where more bridges were built more recently. When we plot the location of each bridge over the map in green, we see that these counties in general have fewer bridges than other counties, which may indicate that they have been developed more recently. There are even some counties which do not include any bridges in the database. People living in these counties may have never even seen a bridge in their lives. 


# Bridge Condition:

```{r}
ggplot(nbi ,  aes(x=yearBuilt, fill = CAT10))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1930, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "All IL Bridge Superstructure Condition by Year Built")

ggplot(subset(nbi, FUNCTIONAL_CLASS_026 %in% (c("01", "11"))),  aes(x=yearBuilt, fill = CAT10))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1930, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "IL Interstates Superstructure Condition by Year Built")

ggplot(subset(nbi, !(FUNCTIONAL_CLASS_026 %in% (c("01", "11")))),  aes(x=yearBuilt, fill = CAT10))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1800, 2020, 20), limits = c(1930, 2020))+
  labs(x= "Year Built", y= "Number of Bridges", title= "IL Non-Interstates Superstructure Condition by Year Built")
```

# Trends

```{r}
nbi_trend$Rating <- factor(nbi_trend$Rating, levels = c("Good", "Fair", "Poor"))

ggplot(nbi_trend ,  aes(x=NBI_Year, fill = FUNCTIONAL_CLASS_026))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1990, 2020, 5), limits = c(1990, 2020))+
  labs(x= "Data Year", y= "Number of Bridges", title= "Number of IL Bridges in Inventory By Class")

ggplot(nbi_trend ,  aes(x=NBI_Year, fill = CAT10))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1990, 2020, 5), limits = c(1990, 2020))+
  labs(x= "Data Year", y= "Number of Bridges", title= "Number of IL Bridges in Inventory By Condition")

ggplot(nbi_trend ,  aes(x=NBI_Year, fill = Rating))+
  geom_bar(position = "stack", show.legend = TRUE) +
  scale_x_continuous(breaks = seq(1990, 2020, 5), limits = c(1990, 2020))+
  labs(x= "Data Year", y= "Number of Bridges", title= "Number of IL Bridges in Inventory By Condition")

nbi_trend$Rating = factor(nbi_trend$Rating, levels = c("Good", "Fair", "Poor"))

ggplot(nbi_trend ,  aes(x=NBI_Year, fill = Rating))+
  geom_bar(position = "fill",  show.legend = TRUE) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(1990, 2020, 5), limits = c(1990, 2020))+
  labs(x= "Data Year", y= "Number of Bridges", title= "Number of IL Bridges in Inventory By Condition")

ggplot(nbi_trend, aes(x=NBI_Year)) +
  geom_bar(aes(y=..count../sum(..count..), fill=Rating)) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(1990, 2020, 5), limits = c(1990, 2020))+
  labs(x= "Data Year", y= "Number of Bridges", title= "Number of IL Bridges in Inventory By Condition")

nbi_trend %>%
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n()) %>%
  mutate(pct = countt/sum(countt)) %>%
  ggplot(aes(x= NBI_Year, y = pct, fill = Rating)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste(format(pct*100, digits = 1), "%")), 
            position = position_stack(vjust = 0.5), size = 2) +
  coord_flip() +
   labs(x= "Year", y= "Percent of Bridges", title= "Percent of IL Bridges in Inventory By Condition")

nbi_trend %>%
  filter(HIGHWAY_SYSTEM_104 == 1) %>%
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n()) %>%
  mutate(pct = countt/sum(countt)) %>%
  ggplot(aes(x= NBI_Year, y = pct, fill = Rating)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste(format(pct*100, digits = 1), "%")), 
            position = position_stack(vjust = 0.5), size = 2) +
  coord_flip() +
   labs(x= "Year", y= "Percent of Bridges", title= "Percent of IL Bridges in Inventory By Condition")

nbi_trend %>% 
  filter(HIGHWAY_SYSTEM_104 == 1) %>%
  filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>%
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n()) %>%
  mutate(pct = countt/sum(countt)) %>%
  ggplot(aes(x= NBI_Year, y = pct, fill = Rating)) + 
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste(format(pct*100, digits = 1), "%")), 
            position = position_stack(vjust = 0.5), size = 2) +
  coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of CMAP Bridges in Inventory By Condition")

nbi_trend %>% 
  filter(HIGHWAY_SYSTEM_104 == 1) %>%
  filter(COUNTY_CODE_003 %in% c("031", "043", "089", "093", "097", "111", "197")) %>%
  filter(Rating != "NA") %>%
  group_by(NBI_Year, Rating) %>%
  summarise(countt = n(), area = sum(area_m)) %>%
  mutate(pct = area/sum(area)) %>%
  ggplot(aes(x= NBI_Year, y = pct*100, fill = Rating)) + 
  geom_bar(stat = "identity") +
  #geom_text(aes(label = paste(format(pct*100, digits = 1), "%")),
  geom_text(aes(label = paste(round(pct*100, digits = 1), "%", sep = "")), 
            position = position_stack(vjust = 0.5), size = 3, angle = 90) +
  scale_x_continuous(breaks = seq(1990, 2020, 2), limits = c(1990, 2020))+
  scale_fill_brewer(type = "seq", palette="Spectral", direction = -1) +
  #coord_flip() +
  labs(x= "Year", y= "Percent of Bridges", title= "Percent of CMAP Bridges Area By Condition") 
  

```


